---
title: "parse_text inside longpca"
output: rmarkdown::html_vignette
editor_options: 
  chunk_output_type: console
vignette: >
  %\VignetteIndexEntry{parse_text}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


This document shows how you can use `parse_text` inside the function `make_interaction_model`.  

```{r setup}
library(dplyr)
library(longpca)
```

Recall that the basic use of `make_interaction_model` in the flights data:

```{r}
library(nycflights13)
formula =  ~ (month & day)*(dest)
im = make_interaction_model(formula, flights)
# pcs = pca(im, k = 6)
```


In this example, the formula is `~ (month & day)*(dest)`.  So, the rows are indexed by the unique values of `month` and `day` and the columns are indexed by the unique values of `dest`. 

```{r}
im$row_universe
im$column_universe
```


If our data contains sentences, or *sequences* of id's (e.g. words/bi-grams/etc), then the "unique values" need to first be parsed/extracted.  For example, loaded into `longpca` is a toy data set `all_packages`.  It is about 20000 rows.  Each row gives information about one R package.

```{r}
# you can get a fresh version here:
# all_packages = available.packages() %>% as_tibble()
all_packages |> select(Package,  Imports)
```

Notice how the column `Imports` gives a "csv" of other packages. We can "extract" these values with `parse_text = TRUE`:
```{r}
im_imports = make_interaction_model(~Package*Imports, tib = all_packages, parse_text= TRUE)
im_imports$row_universe
im_imports$column_universe
```
Under the hood, this calls `tidytext::unnest_tokens` to convert `Imports` into a "bag-of-words".  So, the columns are now indexed by the unique packages that are imported.  There are  20,319 total `Packages` in `all_packages` and each one forms a "row" in `im_imports`.  Only 6,230 of these packages have been imported by another package.  Each of these forms a "column" in `im_imports`.  Here "row" and "column" are in quotes because `im_imports` is not a matrix, but there is a matrix "under the hood" that has these rows and columns.

Additional arguments to `make_interaction_model` are passed off to `tidytext::unest_tokens`.  For example, by default it puts the imported packages into lower case. We can turn that off... 
```{r}
im_imports = make_interaction_model(~Package*Imports, tib = all_packages, parse_text= TRUE, to_lower = FALSE)
im_imports$column_universe
```

Notice how `Rcpp` in row 6 of the `$column_universe` was previously `rcpp` (lower case). You could parse a different variable instead.  For example,

```{r}
im_description = make_interaction_model(~Package*Description, top_packages, parse_text = TRUE, to_lower= TRUE)
diagnose(im_description)
im_description$column_universe
```

Here we see the stopwords.  These will form high degree nodes in your `interaction_model` graph.  A factor that loads heavily on these is often a factor that indicates "document length". 

You can also parse multiple columns with the `&`:

```{r}
im_imports_authors = make_interaction_model(~Package*(Imports&Author&Description&Title), tib = top_packages, parse_text = TRUE)
im_imports_authors$row_universe
im_imports_authors$column_universe

```


So, let's do a quick analysis. We see that the tokens are very very sparse, with 64% of tokens appearing only once! We will want to keep an eye out for localization in the `streaks` plot.

```{r}
diagnose(im_imports_authors)
```

Because it is so sparse, I'm going to use more bootstraps than default in `pick_dim`; 20 should be enough.
```{r}
eigcv = pick_dim(im_imports_authors, num_bootstraps = 20)
plot(eigcv)
eigcv
```
I'm going to pick 10 for convenience. 

```{r}
pcs = pca(im_imports_authors, k = 10)
streaks(pcs)
streaks(pcs, "columns")
```

```{r}
spcs = rotate(pcs)
streaks(spcs)
streaks(spcs, "columns")
```

For example, the second rotated component looks to pick up the cluster of r packages for making html things:
```{r}
top(spcs, 2,keep_how_many = 15)
```
